---
title: "Case_study_Cyclist"
author: "Jade Nguyen"
date: '2022-06-08'
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = FALSE)
```


## 1 Load necessary packages
```{r}
    library(tidyverse)
    library(readr)
    library(janitor)
    library(dplyr)
    library(lubridate)
```

## 2 Collect data
```{r}
    df1 <- read.csv("./data_case_study/202004-divvy-tripdata.csv")
    df2 <- read.csv("./data_case_study/202005-divvy-tripdata.csv")
    df3 <- read.csv("./data_case_study/202006-divvy-tripdata.csv")
    df4 <- read.csv("./data_case_study/202007-divvy-tripdata.csv")
    df5 <- read.csv("./data_case_study/202008-divvy-tripdata.csv")
    df6 <- read.csv("./data_case_study/202009-divvy-tripdata.csv")
    df7 <- read.csv("./data_case_study/202010-divvy-tripdata.csv")
    df8 <- read.csv("./data_case_study/202011-divvy-tripdata.csv")
    df9 <- read.csv("./data_case_study/202012-divvy-tripdata.csv")
    df10 <- read.csv("./data_case_study/202101-divvy-tripdata.csv")
    df11 <- read.csv("./data_case_study/202102-divvy-tripdata.csv")
    df12 <- read.csv("./data_case_study/202103-divvy-tripdata.csv")
    df13 <- read.csv("./data_case_study/202104-divvy-tripdata.csv")
    df14 <- read.csv("./data_case_study/202105-divvy-tripdata.csv")
```

## 3 Combine data
```{r}
    trip_data <- rbind(df1,df2,df3,df4,df5,df6,df7,df8,df9,df10,df11,df12,df13,df14)
```

## 4 Add columns for date, month, day and year of each ride
```{r}
    trip_data$date <- as.Date(trip_data$started_at)
    trip_data$year <- format(as.Date(trip_data$date), "%Y") 
    trip_data$month <- format(as.Date(trip_data$date), "%m") 
    trip_data$day <- format(as.Date(trip_data$date), "%d")
    trip_data$day_of_week <- format(as.Date(trip_data$date), "%A")
```

## 5 Add ride_length calculation in minutes and convert to number
```{r}
    trip_data$ride_length <- difftime(trip_data$ended_at,trip_data$started_at, units = "mins")
    trip_data$ride_length <- as.numeric(trip_data$ride_length)
```

## 6 Get summary of ride_length data
```{r}
    summary(trip_data$ride_length)
```
The min value is a negative number.
So next, I use filter to find out how many negative data exist
```{r}
    trip_data %>% filter(ride_length < 0) 
```
Result shows 10,559 rows of negative data.
To carry on with analysis, I decided to remove the negative data


## 7 Remove negative ride length and na value 
```{r}
    df <- trip_data %>%  filter(ride_length > 0) %>% drop_na()
```
This reduces data from 4358611 rows to 4211741

## 8 Find number of casual riders and number of member riders
```{r}
    count(df,member_casual)
```
###      Export data for future visualization
```{r}
    data_viz_1 <- count(df,member_casual)
    write.csv(data_viz_1, "data_viz_1.csv")
```

## 9 Summarize data for each group of riders
```{r}
   df %>% 
    group_by(member_casual) %>% 
    summarise(avg_ride_length = mean(ride_length),
              min_ride_length= min(ride_length),
              max_ride_length= max(ride_length))
```
###      Export data for future visualization
```{r}
  data_viz_2 <-  df %>% 
    group_by(member_casual) %>% 
    summarise(avg_ride_length = mean(ride_length),
              min_ride_length= min(ride_length),
              max_ride_length= max(ride_length))
  write.csv(data_viz_2, "data_viz_2.csv")
```
## 10 Rides by member_casual and weekday
```{r}
    df %>% 
      mutate(weekday = wday(started_at, label = TRUE)) %>%
      group_by(member_casual, weekday) %>%
      summarise(number_of_rides = n()	 
                ,average_ride_length = mean(ride_length)) %>%
      arrange(member_casual, weekday)
```
###      Export data for future visualization
```{r}
    data_viz_3 <- df %>% 
                mutate(weekday = wday(started_at, label = TRUE)) %>%
                group_by(member_casual, weekday) %>%
                summarise(number_of_rides = n()	 
                          ,average_ride_length = mean(ride_length)) %>%
                arrange(member_casual, weekday)
    write.csv(data_viz_3, "data_viz_3.csv")  
```
## 11 Rides by member_casual and month
```{r}
   df %>%
    group_by(member_casual, year, month) %>%
    summarise(number_of_rides = n()	 
              ,average_ride_length = mean(ride_length)) %>%
    arrange(member_casual, month) %>% 
    tbl_df %>% print(n = Inf) 
```

###      Export data for future visualization
#### For this visualization I remove data from 4/2021 and 5/2021 in excel file before visualizing it in Tableau

    install.packages("openxlsx")
    library(openxlsx)
    data_viz_4 <- df %>%
                group_by(member_casual, year, month) %>%
                summarise(number_of_rides = n()	 
                          ,average_ride_length = mean(ride_length)) %>%
                arrange(member_casual, month)
       
    write.xlsx(data_viz_4, 'data_viz_4.xlsx')


# Refer to [Tableau Viz](https://public.tableau.com/views/PortfolioProject_Cyclist/Dashboard1?:language=en-US&:display_count=n&:origin=viz_share_link) to see data vizs